---
title: "Body size analysis"
author: "Martin Garlovsky"
date: "2022-08-04"
output: 
  workflowr::wflow_html:
    code_folding: hide 
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

#### Load packages
```{r}
library(tidyverse)

library(lme4)
library(DHARMa)
library(emmeans)
library(kableExtra)
library(knitrhooks) # install with devtools::install_github("nathaneastwood/knitrhooks")
library(showtext)
output_max_height() # a knitrhook option

options(stringsAsFactors = FALSE)

# colour palettes
met.pal <- MetBrewer::met.brewer('Johnson')
met3 <- met.pal[c(1, 3, 5)]

# set contrasts
options(contrasts = c("contr.sum", "contr.poly"))

```

#### Load data
```{r}
bd_dat <- read.csv('data//wrangled/body_data.csv') %>% 
  mutate(mito_snp = as.factor(mito_snp),
         coevolved = if_else(mito == nuclear, "matched", "mismatched"))
```

### Geography based analysis {.tabset}

We collected 4 offspring of each sex from 4 females - a family (`fam`) in our analysis - and measured landmarks on the wing as a proxy for body size. We modelled body size with a linear mixed model, with scaled wing area as the response. 

#### Look at the raw data 
```{r}
bd_dat %>% 
  ggplot(aes(x = nuclear, y = scaled_area, fill = mito)) +
  geom_boxplot() +
  facet_wrap(~sex, scales = 'free_x') +
  scale_colour_manual(values = met3) +
  scale_fill_manual(values = met3) +
  theme_bw() + 
  theme() + 
  NULL
```


```{r}
hist(bd_dat$scaled_area, breaks = 50)

b1full <- lmerTest::lmer(scaled_area ~ mito * nuclear * sex + (sex|LINE) + (1|fam), 
                         data = bd_dat, REML = TRUE)
```

#### Check model diagnostics
```{r, fig.height=12}
performance::check_model(b1full)
```

### Results {.tabset}
```{r}
anova(b1full, type = "III", ddf = "Kenward-Roger") %>% broom::tidy() %>% 
  as_tibble() %>% 
  kable(digits = 3, 
        caption = 'Type III Analysis of Variance Table with Kenward-Roger`s method') %>% 
  kable_styling(full_width = FALSE)

#summary(b1full)

bind_rows(emmeans(b1full, pairwise ~ sex, adjust = "tukey")$contrasts %>% as_tibble(),
          emmeans(b1full, pairwise ~ nuclear, adjust = "tukey")$contrasts %>% as_tibble()) %>% 
  kable(digits = 3, 
        caption = 'Posthoc Tukey tests to compare which groups differ') %>% 
  kable_styling(full_width = FALSE) %>%
  kableExtra::group_rows("Sex", 1, 1) %>%
  kableExtra::group_rows("Nuclear", 2, 4)

```

#### Reaction norms
```{r}
body2_emm <- emmeans(b1full, ~ mito * nuclear * sex)

# reaction norms
body_norms <- emmeans(body2_emm, ~ mito * nuclear * sex, type = 'response') %>% as_tibble() %>% 
  ggplot(aes(x = nuclear, y = emmean, fill = mito)) +
  geom_line(aes(group = mito, colour = mito), position = position_dodge(width = .5)) +
  geom_point(size = 3, pch = 21, position = position_dodge(width = .5)) +
  labs(y = 'Scaled wing area (EMM)') +
  scale_colour_manual(values = met3) +
  scale_fill_manual(values = met3) +
  facet_wrap(~ sex, labeller = as_labeller(c(f = "Female", m = "Male"))) +
  theme_bw() + 
  theme()

body_norms

```

#### Raw data with means
```{r}
body_size <- emmeans(body2_emm, ~ mito * nuclear * sex, type = 'response') %>% as_tibble() %>% 
  ggplot(aes(x = nuclear, y = emmean, fill = mito)) +
  geom_jitter(data = bd_dat,
              aes(y = scaled_area, colour = mito),
               position = position_jitterdodge(dodge.width = .5, jitter.width = .1),
              size = 0.75, alpha = .1) +
  geom_jitter(data = bd_dat %>% 
                group_by(LINE, sex) %>% 
                summarise(mn = mean(scaled_area)) %>% 
                separate(LINE, into = c("mito", "nuclear", NA), sep = "(?<=.)", remove = FALSE),
              aes(y = mn, colour = mito),
              position = position_jitterdodge(dodge.width = .5, jitter.width = .1),
              alpha = .85, size = 2) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
                width = .25, position = position_dodge(width = .5)) +
  geom_point(size = 3, pch = 21, position = position_dodge(width = .5)) +
  labs(y = 'Scaled wing area (EMM ± 95% CIs)') +
  scale_colour_manual(values = met3) +
  scale_fill_manual(values = met3) +
  facet_wrap(~ sex, labeller = as_labeller(c(f = "Female", m = "Male"))) +
  theme_bw() + 
  theme() + 
  geom_text(data = bd_dat %>% group_by(mito, nuclear, sex) %>% count(), aes(y = -2.2, label = n), 
            size = 2, position = position_dodge(width = .5)) + 
  NULL

body_size

```

## Matched vs. mismatched
```{r}
body_matched_lmerTest <- lmerTest::lmer(scaled_area ~ sex * coevolved + (sex|LINE) + (1|fam), 
                                        data = bd_dat, REML = TRUE)

anova(body_matched_lmerTest, type = "III", ddf = "Kenward-Roger")
#summary(body_matched_lmerTest)

```


## Mito-type analysis
```{r}
snp_full <- lmerTest::lmer(scaled_area ~ mito_snp * nuclear * sex + (sex|LINE) + (1|fam), 
                           data = bd_dat,
                           na.action = 'na.fail', REML = TRUE)

anova(snp_full, type = "III", ddf = "Kenward-Roger")

# snp_full_emm <- emmeans(snp_full, ~ mito_snp * nuclear * sex)
# pairs(snp_full_emm, simple = "each")

emmeans(snp_full, ~ mito_snp * nuclear * sex) %>% as_tibble() %>% drop_na() %>% 
  ggplot(aes(x = nuclear, y = emmean, fill = mito_snp)) +
  geom_jitter(data = bd_dat,
              aes(y = scaled_area, colour = mito_snp),
               position = position_jitterdodge(dodge.width = .5, jitter.width = .1),
              alpha = .25) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
                width = .25, position = position_dodge(width = .5)) +
  geom_point(size = 3, pch = 21, position = position_dodge(width = .5)) +
  labs(y = 'Scaled wing area (EMM ± 95% CIs)') +
  scale_colour_viridis_d(option = "H") +
  scale_fill_viridis_d(option = "H") +
  facet_wrap(~sex, labeller = as_labeller(c(f = "Female", m = "Male"))) +
  theme_bw() + 
  theme() +
  #ggsave('figures/body_size_mitocluster.pdf', height = 3.5, width = 7.5, dpi = 600, useDingbats = FALSE) +
  NULL

```
