---
title: "Female fertility"
author: "Martin Garlovsky"
date: "2022-08-03"
output: 
  workflowr::wflow_html:
    code_folding: hide 
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

#### Load packages
```{r}
library(tidyverse)

library(lme4)
#library(merTools)
library(DHARMa)
library(emmeans)
library(kableExtra)
library(knitrhooks) # install with devtools::install_github("nathaneastwood/knitrhooks")
library(showtext)

library(conflicted)

select <- dplyr::select
filter <- dplyr::filter

output_max_height() # a knitrhook option

options(stringsAsFactors = FALSE)

# colour palettes
met.pal <- MetBrewer::met.brewer('Johnson')
met3 <- met.pal[c(1, 3, 5)]

# set contrasts
options(contrasts = c("contr.sum", "contr.poly"))

```

#### Load data
```{r}
fert_dat <- read_csv("data/wrangled/female_fertility.csv") %>% 
  mutate(mito_snp = as.factor(mito_snp),
         coevolved = if_else(mito == nuclear, "matched", "mismatched"))
```

# Introduction
First look at the data shows that fecundity appears to plateau between the first and second vial before declining. Therefore, we model fecundity across three episodes: 

1. early life fecundity
2. lifetime fecundity
3. the rate of progeny production

```{r}
# progeny by mito / nuclear
fert_dat %>% 
  group_by(mito, nuclear, vial) %>% 
  summarise(mn = mean(progeny),
            se = sd(progeny)/sqrt(n()),
            s95 = se * 1.96) %>% 
  ggplot(aes(x = vial, y = mn, colour = mito)) +
  geom_point(data = fert_dat,
             aes(y = progeny, colour = mito), alpha = .15,
             position = position_jitterdodge(dodge.width = .5, jitter.width = .1)) +
  geom_ribbon(aes(ymin = mn - s95, ymax = mn + s95, fill = mito, group = mito),
              alpha = .5, colour = NA) +
  geom_errorbar(aes(ymin = mn - s95, ymax = mn + s95), 
                width = .25, position = position_dodge(width = .5)) +
  geom_point(position = position_dodge(width = .5)) +
  geom_line(aes(group = nuclear)) + 
  facet_wrap(~nuclear) +
  scale_x_continuous(breaks = c(1:7)) +
  scale_colour_manual(values = met3) +
  scale_fill_manual(values = met3) +
  labs(y = 'Fecundity per vial (EMM ± 95% CIs)') +
  theme_bw() + 
  theme() + 
  #ggsave('figures/vial_means.pdf', height = 3, width = 9, dpi = 600, useDingbats = FALSE) +
  NULL

```
**Figure 1.** Per vial progeny production for each mitonculear genotype. Facets are nuclear genotypes, with colour denoting mitochondrial genotype. 

# > Early life fecundity
We summed progeny produced from the first two vial for each female. 
```{r}
daydat <- fert_dat %>% filter(vial == '1' | vial == '2') %>% 
  dplyr::select(ID:mtn, mito_snp, coevolved, LINE, vial, progeny) %>% 
  pivot_wider(names_from = vial,
              values_from = progeny) %>% 
  dplyr::rename(v1 = `1`, v2 = `2`) %>% 
  mutate(comb_vial = rowSums(across(c(v1, v2))))

daydat %>% pivot_longer(cols = c(v1, v2)) %>% #filter(age == "young") %>% 
  ggplot(aes(x = name, y = value)) + 
  geom_line(aes(group = ID, colour = mito)) +
  tidybayes::stat_halfeye(alpha = .5) +
  scale_colour_manual(values = met3) +
  labs(y = 'Female fecundity') +
  facet_grid(mito ~ nuclear) +
  NULL
```
**Figure 2.** Female fecundity in vial 1 and vial 2. Facets are for mito and nuclear genotypes. Lines connect individual females with large points representing the mean and error bars summarise the 66% and 95% quantiles.

```{r}
# combined vial 1 and vial 2
#hist(daydat$comb_vial)
```


```{r}
fecundity_early <- lmerTest::lmer(comb_vial ~ mito * nuclear + (1|LINE), data = daydat, REML = TRUE)
```

### >>> Check diagnostics
```{r, fig.height=12}
performance::check_model(fecundity_early)
```


```{r}
anova(fecundity_early, type = "III", ddf = "Kenward-Roger") %>% broom::tidy() %>% 
  as_tibble() %>% 
  kable(digits = 3, 
        caption = 'Type III Analysis of Variance Table with Kenward-Roger`s method') %>% 
  kable_styling(full_width = FALSE)

emmeans(fecundity_early, pairwise ~ nuclear, adjust = "tukey")$contrasts %>% broom::tidy() %>% 
  kable(digits = 3, 
        caption = 'Posthoc Tukey tests for days effect. Results are averaged over the levels of mito') %>% 
  kable_styling(full_width = FALSE)

```

### >>> Matched vs. mismatched
```{r}
coevo_early <- lmerTest::lmer(comb_vial ~ coevolved + (1|LINE), data = daydat, REML = TRUE)

anova(coevo_early, type = "III", ddf = "Kenward-Roger")

```

### >>> Mito-type analysis
```{r}
fecundity_day1_snp <- lmerTest::lmer(comb_vial ~ mito_snp * nuclear + (1|LINE), data = daydat, REML = TRUE)
anova(fecundity_day1_snp, type = "III", ddf = "Kenward-Roger")

early_fec_snp <- emmeans(fecundity_day1_snp, ~ mito_snp * nuclear, type = 'response') %>% 
  as_tibble() %>% drop_na() %>% 
  ggplot(aes(x = nuclear, y = emmean, fill = mito_snp)) +
  geom_jitter(data = daydat,
              aes(y = comb_vial, colour = mito_snp),
               position = position_jitterdodge(dodge.width = .5, jitter.width = .1),
              alpha = .25) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
                width = .25, position = position_dodge(width = .5)) +
  geom_point(size = 3, pch = 21, position = position_dodge(width = .5)) +
  labs(y = 'Early life fecundity (EMM ± 95% CIs)') +
  scale_colour_viridis_d(option = "H") +
  scale_fill_viridis_d(option = "H") +
  theme_bw() + 
  theme() + 
  NULL

early_fec_snp

```


# > Lifetime fecundity

Here we sum the total number of offspring each female produced across the entire 21 days.
```{r}
# summarised data 
sum_fecundity <- fert_dat %>% 
  group_by(ID, mito, mito_snp, mtgrp, nuclear, mtn, coevolved, LINE) %>% 
  summarise(total.fec = sum(progeny)) %>% 
  ungroup() %>% 
  mutate(# scale variables 
    scaled_fec = as.numeric(scale(total.fec)),
    # add observation level random effect
    OLRE = 1:nrow(.))

# check mtgrp not crossed within lines
#xtabs(~ LINE + mtgrp, data = sum_fecundity)

hist(sum_fecundity$total.fec, breaks = 50)

# fit linear fit
fecundity_total <- lmerTest::lmer(total.fec ~ mito * nuclear + (1|LINE), data = sum_fecundity, REML = TRUE)
```

### >>> Check diagnostics
```{r, fig.height=12}
performance::check_model(fecundity_total)
```


```{r}
anova(fecundity_total, type = "III", ddf = "Kenward-Roger")

# grand mean
#emmeans::emmeans(fecundity_total, specs = ~1, type = "response")

```

### >>> Matched vs. mismatched
```{r}
fecundity_coevo <- lmerTest::lmer(total.fec ~ coevolved + (1|LINE), data = sum_fecundity, REML = TRUE)
anova(fecundity_coevo, type = "III", ddf = "Kenward-Roger")

```

### >>> Mito-type analysis
```{r}
fecundity_lmer_snp <- lmerTest::lmer(total.fec ~ mito_snp * nuclear + (1|LINE), data = sum_fecundity)
anova(fecundity_lmer_snp, type = "III", ddf = "Kenward-Roger")

lifetime_fec_snp <- emmeans(fecundity_lmer_snp, ~ mito_snp * nuclear) %>% as_tibble() %>% drop_na() %>% 
  ggplot(aes(x = nuclear, y = emmean, fill = mito_snp)) +
  geom_jitter(data = sum_fecundity,
              aes(y = total.fec, colour = mito_snp),
               position = position_jitterdodge(dodge.width = .5, jitter.width = .1),
              alpha = .5) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
                width = .25, position = position_dodge(width = .5)) +
  geom_point(size = 3, pch = 21, position = position_dodge(width = .5)) +
  labs(y = 'Lifetime fecundity (EMM ± 95% CIs)',
       colour = 'nuclear', fill = 'nuclear') +
  scale_colour_viridis_d(option = "H") +
  scale_fill_viridis_d(option = "H") +
  theme_bw() +
  #theme(legend.position = 'bottom') +
  NULL

lifetime_fec_snp

```


# > Rate of decline {.tabset}
We modelled the rate of progeny production using a GLMM during the period of decline, namely from vial 2 to 6 (excluding vials 1 and 7). A quick plot of the data suggests a non-linear relationship between progeny production and vial. 

```{r}
fert_filt <- fert_dat %>% filter(vial != '1', vial <= 6)
```

## >> Linear
```{r}
fert_filt %>% 
  ggplot(aes(x = vial, y = progeny, colour = mito)) +
  geom_point(alpha = .25, position = position_dodge(width = .5)) +
  stat_smooth(method = "lm", formula = y ~ poly(x, 1), colour = "black") +
  facet_wrap(~ nuclear) +
  scale_x_continuous(breaks = c(1:7)) +
  scale_colour_manual(values = met3) +
  scale_fill_manual(values = met3) +
  theme_bw() + 
  theme() + 
  NULL
```

## >> Polynomial
```{r}
fert_filt %>% 
  ggplot(aes(x = vial, y = progeny, colour = mito)) +
  geom_point(alpha = .25, position = position_dodge(width = .5)) +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2), colour = "black") +
  facet_wrap(~ nuclear) +
  scale_x_continuous(breaks = c(1:7)) +
  scale_colour_manual(values = met3) +
  scale_fill_manual(values = met3) +
  theme_bw() + 
  theme() + 
  NULL

```

#

## >> Fit the model
We compared a linear fit to a polynomial fit. Based on AICc, the second order polynomial fit is preferred. 

```{r}
# fit model
rate_glmm <- glmer(progeny ~ mito * nuclear * vial + (1|LINE:vial) + (1|ID) + (1|OLRE),
                   data = fert_filt, family = 'poisson',
                   control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 50000)))

# fit model
rate_poly <- glmer(progeny ~ mito * nuclear * vial + I(vial^2) + (1|LINE:vial) + (1|ID) + (1|OLRE),
                   data = fert_filt, family = 'poisson',
                   control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 50000)))

MuMIn::AICc(rate_glmm, rate_poly)

```

### >>> Check diagnostics
```{r, fig.height=12}
performance::check_model(rate_poly)
performance::check_overdispersion(rate_poly) # not overdispersed
performance::check_zeroinflation(rate_poly) # zero inflated

```

```{r}

testDispersion(rate_poly)
simulationOutput <- simulateResiduals(fittedModel = rate_poly, plot = FALSE)
testZeroInflation(simulationOutput)
hist(residuals(simulationOutput))
hist(residuals(simulationOutput, quantileFunction = qnorm, outlierValues = c(-7,7)))
plot(simulationOutput)

```

## >> Results
```{r}
car::Anova(rate_poly, type = 'III')
#summary(rate_glmm)

emt <- emtrends(rate_poly, "nuclear", var = "vial")
#pairs(emt)

rate_norms <- emtrends(rate_poly, c("mito", "nuclear"), var = "vial") %>% as_tibble() %>% 
  ggplot(aes(x = nuclear, y = vial.trend, fill = mito)) +
  geom_line(aes(group = mito, colour = mito), position = position_dodge(width = .5)) +
  geom_point(size = 3, pch = 21, position = position_dodge(width = .5)) +
  labs(y = 'Fecundity (EMM)') +
  scale_colour_manual(values = met3) +
  scale_fill_manual(values = met3) +
  theme_bw() + 
  theme() + 
  NULL

```

##### >>>>> Model predicted values
Here we generate model predicted values for a model with a simplified random effects structure. This is unused in the manuscript at present. 
```{r}
# reduced model for predicted values
# # plot data and model fit
rate_glmm_dummy <- glmer(progeny ~ mito * nuclear * vial + I(vial^2) + (1 + vial|LINE),
                         data = fert_filt, family = 'poisson',
                         control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 50000)))

#car::Anova(rate_glmm_dummy, type = 'III')

prediction_data <- expand.grid(vial = seq(from = 2, to = 6, length = 100), 
                               #ID = unique(fert_filt$ID),
                               #OLRE = unique(fert_filt$OLRE),
                               LINE = levels(as.factor(fert_filt$LINE))) %>%
  mutate(mito = str_sub(LINE, 1, 1),
         nuclear = str_sub(LINE, 2, 2))

mySumm <- function(.) {
  predict(., newdata=prediction_data, re.form=NULL)
}
####Collapse bootstrap into median, 95% PI
sumBoot <- function(merBoot) {
  return(
    data.frame(fit = apply(merBoot$t, 2, function(x) as.numeric(quantile(x, probs=.5, na.rm=TRUE))),
               lwr = apply(merBoot$t, 2, function(x) as.numeric(quantile(x, probs=.025, na.rm=TRUE))),
               upr = apply(merBoot$t, 2, function(x) as.numeric(quantile(x, probs=.975, na.rm=TRUE)))
    )
  )
}

### Reader beware - this takes a long time to run!!!
##lme4::bootMer() method 2
# PI.boot2.time <- system.time(
#   boot2 <- lme4::bootMer(rate_glmm_dummy, mySumm, nsim=1000, use.u=TRUE, type="parametric", .progress = "txt")
# )
##saveRDS(boot2, file = "output/female_rates_poly.Rdata")

boot2 <- read_rds("output/female_rates_poly.Rdata")
PI.boot2 <- sumBoot(boot2)

plot_data <- data.frame(prediction_data, PI.boot2)
#head(plot_data)
colnames(plot_data)[5] <- "progeny"

plot_data_summary <- plot_data %>% 
  group_by(nuclear, vial) %>% 
  summarise(mn_fit = mean(progeny),
            #mn_sef = mean(se.fit),
            mn_lwr = mean(lwr),
            mn_upr = mean(upr)
            ) %>% 
  ungroup() %>% mutate(across(3:5, ~exp(.x)))

# plot
fert_mns <- fert_filt %>% 
  group_by(nuclear, vial) %>% 
  summarise(mn = mean(progeny),
            se = sd(progeny)/sqrt(n()),
            s95 = se * 1.96)

rateP <- fert_filt %>% 
  ggplot(aes(x = vial, y = progeny, colour = nuclear)) +
  geom_point(aes(y = progeny, colour = nuclear, shape = mito), alpha = .15,
             position = position_jitterdodge(dodge.width = .5, jitter.width = .1)) +
  geom_ribbon(data = plot_data_summary,
              aes(y = mn_fit, ymin = mn_lwr, ymax = mn_upr, fill = nuclear, group = nuclear),
              alpha = .5, colour = NA) +
  geom_line(data = plot_data_summary,
            aes(y = mn_fit, colour = nuclear, group = nuclear)) +
  geom_errorbar(data = fert_mns,
                aes(y = mn, ymin = mn - s95, ymax = mn + s95),
                width = .25, position = position_dodge(width = .5)) +
  geom_point(data = fert_mns, aes(y = mn), position = position_dodge(width = .5)) +
  scale_x_continuous(breaks = c(1:7)) +
  scale_colour_manual(values = met3) +
  scale_fill_manual(values = met3) +
  labs(y = 'Fecundity per vial (mean ± SE)') +
  theme_bw() + 
  theme() + 
  #ggsave('figures/vial_means.pdf', height = 3, width = 9, dpi = 600, useDingbats = FALSE) +
  NULL

rateP

# raw data
fem_fert_rawplot <- fert_dat %>% 
  group_by(mito, nuclear, vial) %>% 
  summarise(mn = mean(progeny),
            se = sd(progeny)/sqrt(n()),
            s95 = se * 1.96) %>% 
  ggplot(aes(x = vial, y = mn, colour = mito)) +
  geom_point(data = fert_dat,
             aes(y = progeny, colour = mito), alpha = .15,
             position = position_jitterdodge(dodge.width = .5, jitter.width = .1)) +
  geom_errorbar(aes(ymin = mn - s95, ymax = mn + s95), 
                width = .25, position = position_dodge(width = .5)) +
  geom_point(position = position_dodge(width = .5)) +
  scale_x_continuous(breaks = c(1:7)) +
  scale_colour_manual(values = met3) +
  scale_fill_manual(values = met3) +
  labs(y = 'Fecundity per vial (mean ± SE)') +
  facet_wrap(~ nuclear) +
  theme_bw() + 
  theme() + 
  NULL
#ggsave(filename = 'figures/fem_fert_rawplot.pdf', height = 4, width = 9, dpi = 600, useDingbats = FALSE)

```

### >>> Matched vs. mismatched
```{r}

rate_coevo <- glmer(progeny ~ coevolved * vial + (1|LINE:vial) + (1|ID) + (1|OLRE),
                         data = fert_filt, family = 'poisson',
                         control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 50000)))

car::Anova(rate_coevo, type = 'III')

```

### >>> Mito-type analysis
```{r}
# fit model
rate_glmm_snp <- glmer(progeny ~ mito_snp * nuclear * vial + (1|LINE:vial) + (1|ID) + (1|OLRE),
                       data = fert_filt, family = 'poisson', 
                       control= glmerControl(optimizer="bobyqa", optCtrl=list(maxfun=50000)))

car::Anova(rate_glmm_snp, type = 'III')
#summary(rate_glmm_snp)

```

# > Combined plot for all stages
```{r}
comb_all_female <- bind_rows(
  emmeans(fecundity_early, ~ mito * nuclear) %>% as_tibble() %>% mutate(stage = "a) Early life"),
  emmeans(fecundity_total, ~ mito * nuclear) %>% as_tibble() %>% mutate(stage = "b) Lifetime"),
  emtrends(rate_glmm, c("mito", "nuclear"), var = "vial") %>% as_tibble() %>% 
    mutate(stage = "c) Rate of decline",
           emmean = vial.trend)) %>% 
  ggplot(aes(x = nuclear, y = emmean, fill = mito)) +
  geom_line(aes(group = mito, colour = mito), position = position_dodge(width = .5)) +
  geom_point(size = 3, pch = 21, position = position_dodge(width = .5)) +
  labs(y = 'Fecundity (EMM)') +
  scale_colour_manual(values = met3) +
  scale_fill_manual(values = met3) +
  facet_wrap(~stage, scales = "free_y") +
  theme_bw() + 
  theme() + 
  NULL

comb_all_female
#ggsave(filename = 'figures/femaleplot_norms3.pdf', height = 4, width = 12, dpi = 600, useDingbats = FALSE)

```


# > Survival analysis

Finally, we modelled reproductive senescence using survival analysis based on the onset of infertility, i.e., the time (in vials) to event (final vial producing progeny for each female) as the response. 

```{r}
library(coxme)
library(survminer)

fsurv <- fert_dat %>% 
  mutate(bin_prog = if_else(progeny == 0, 0, 1),
         mito_snp = as.factor(mito_snp))

fsurv$cum_pr <- ave(fsurv$progeny, fsurv$ID, FUN = cumsum)
fsurv <- fsurv %>% group_by(ID) %>% slice(which.min(progeny)) %>% mutate(EVENT = 1)

fit1 <- survfit(Surv(vial, EVENT) ~ mito + nuclear, data = fsurv)
#summary(fit1)
#anova(fit1)

survplot <- ggsurvplot(fit1, colour = "mito",
                       palette = rep(rev(met3), 3))

survplot$plot + 
  facet_wrap(~nuclear) + 
  theme_bw() + 
  theme(legend.position = "") +
  NULL

```

### >>> Check diagnostics
Using a simplified model without the random effects we can check the proportional hazards assumption.
```{r}
plot(survfit(Surv(vial, EVENT) ~ mito + nuclear, data = fsurv), 
     lty = 1:4, 
     fun="cloglog", main = "cloglog")

test.ph <- cox.zph(coxph(Surv(vial, EVENT) ~ mito * nuclear, data = fsurv))

ggcoxzph(test.ph)

```

We then use likelihood ratio tests comparing the full model to submodels excluding the effect of interest. 
```{r}
cox.des <- coxme(Surv(vial, EVENT) ~ mito * nuclear + (1|LINE), data = fsurv)

cox.des.n1 <- coxme(Surv(vial, EVENT) ~ 1 + nuclear + (1|LINE), data = fsurv)
cox.des.n2 <- coxme(Surv(vial, EVENT) ~ mito + 1 + (1|LINE), data = fsurv)
cox.des.n3 <- coxme(Surv(vial, EVENT) ~ mito + nuclear + (1|LINE), data = fsurv)

# posthoc tests
bind_rows(anova(cox.des, cox.des.n1) %>% broom::tidy(),
          anova(cox.des, cox.des.n2) %>% broom::tidy(),
          anova(cox.des, cox.des.n3) %>% broom::tidy()) %>% #drop_na() %>% 
  kable(digits = 3, 
        caption = 'Likelihood ratio tests comparing the full model to submodels excluding the main effect of interest') %>% 
  kable_styling(full_width = FALSE) %>%
  kableExtra::group_rows("Mitochondria", 1, 2) %>%
  kableExtra::group_rows("Nuclear", 3, 4) %>%
  kableExtra::group_rows("Mitonuclear interaction", 5, 6)


cox_emm <- emmeans(cox.des, ~ mito * nuclear)

# posthoc tests
emmeans(cox_emm, pairwise ~ nuclear, adjust = "tukey")
#emmeans(cox_emm, pairwise ~ nuclear, adjust = "tukey", type = "response")

# posthoc tests
bind_rows(emmeans(cox_emm, pairwise ~ nuclear, adjust = "tukey")$contrasts %>% broom::tidy(),
          emmeans(cox_emm, pairwise ~ nuclear, adjust = "tukey", type = "response")$contrasts %>% broom::tidy()) %>% 
  kable(digits = 3, 
        caption = 'Posthoc Tukey tests to compare which groups differ. Results are averaged over the levels of mito') %>% 
  kable_styling(full_width = FALSE) %>%
  kableExtra::group_rows("log scale", 1, 3) %>%
  kableExtra::group_rows("Hazard ratios", 4, 6)

```

### >>> Matched vs. mismatched
```{r}
cox.coevo <- coxme(Surv(vial, EVENT) ~ coevolved + (1|LINE), data = fsurv)

#anova(cox.coevo)
summary(cox.coevo)

cox.coevo.n1 <- coxme(Surv(vial, EVENT) ~ 1 + (1|LINE), data = fsurv)

anova(cox.coevo, cox.coevo.n1) # mito NS

```

### >>> Mito-type analysis
The mito-type analysis throws an error due to the rank deficiency of the design. 
```{r}

# cox.snp <- coxme(Surv(vial, EVENT) ~ mito_snp * nuclear + (1|LINE), data = fsurv)
# 
# summary(cox.snp)
# 
# cox.snp.n1 <- coxme(Surv(vial, EVENT) ~ 1 + nuclear + (1|LINE), data = fsurv)
# cox.snp.n2 <- coxme(Surv(vial, EVENT) ~ mito_snp + 1 + (1|LINE), data = fsurv)
# cox.snp.n3 <- coxme(Surv(vial, EVENT) ~ mito_snp + nuclear + (1|LINE), data = fsurv)
# 
# anova(cox.snp, cox.snp.n1) # mito NS
# anova(cox.snp, cox.snp.n2) # nuclear sig
# anova(cox.snp, cox.snp.n3) # interaction NS

```
